Fractional Types

### 2019-11-22 fracGC/p.tex
fix centering
code for examples
fix small typo

%% for Sec. 6
\input{latex/J.tex}
\input{latex/PiPointedFracCode.tex}
\input{latex/ExtractionCode.tex}
\input{latex/BooleanCircuitsCode.tex}

@@

\paragraph*{Outline.} We solve the ancilla problem in reversible and
quantum computation using a novel concept: \emph{fractional types}. In
the next section, we introduce the problem of ancilla, motivate its
importance, and explain the limitations of current solutions. Inm
Sec.~\ref{sec:pi}, we review existing work%
~\cite{rc2011,DBLP:conf/esop/CaretteS16,rc2012,James:2012:IE:2103656.2103667}
that introduces a reversible programming language built using type
isomorphism.  Although the concept of fractional types might apply to
general-purpose languages (after some adaptation), its natural
technical definition exploits symmetries present in the categorical
model of type isomorphisms. We present such a definition in
Sec.~\ref{sec:dyn} which captures and illustrates the main novelties
of fractional types and their runtime representations as first-class
gc processes. This definition allows allocation and de-allocation of
ancillae in patterns beyond the scope model but, like existing
stack-based solutions, still requires a runtime check to verify the
safety of de-allocation.
%% \ednote{JC: I don't understand the previous sentence. What does
%% ``this presentation'' refer to?  The paper as a whole, or the
%% details of the fractional types? There is also a tension between
%% ``partially solves'' and ``more general and more expressive'' which
%% might very well be true, but makes for a very vague sentence.}
To make the arguments regarding memory
management explicit, Sec.~\ref{sec:space} presents an abstract machine
with a heap component that grows and shrinks as ancillae are allocated
and de-allocated via garbage collection. In Sec.~\ref{sec:dep} we
lift ``plain'' programs manipulating ancillae to the type system in
order to exploit the power of dependent types to guarantee, for the
first time, the condition necessary for safe de-allocation of
ancillae. Technically this solution requires the introduction of
pointed types on top of which are defined both a monad and comonad of
singleton types. The tracking of ancillae value is done within the
type system via the singleton types. A number of examples exploiting
the singleton-based type system are presented to demonstrate the full
expressiveness of reasoning about fractional types. The type safety proofs demonstrate
that plain (but guaranteed safe)
programs can be extracted and executed without errors or runtime
checks in the conventional type system. The last section concludes
with a discussion of related work and a summary of our results.

@@

\centering{
\input{revrev.tikz}
}

@@

>>>>

%% for Sec. 6
\input{latex/J.tex}
\input{latex/PiPointedFracCode.tex}
\input{latex/ExtractionCode.tex}
\input{latex/BooleanCircuitsCode.tex}
\input{latex/PiPointedFracTraceCode.tex}

@@

\paragraph*{Outline.} We solve the ancilla problem in reversible and
quantum computation using a novel concept: \emph{fractional types}. In
the next section, we introduce the problem of ancilla, motivate its
importance, and explain the limitations of current solutions. In
Sec.~\ref{sec:pi}, we review existing work%
~\cite{rc2011,DBLP:conf/esop/CaretteS16,rc2012,James:2012:IE:2103656.2103667}
that introduces a reversible programming language built using type
isomorphism.  Although the concept of fractional types might apply to
general-purpose languages (after some adaptation), its natural
technical definition exploits symmetries present in the categorical
model of type isomorphisms. We present such a definition in
Sec.~\ref{sec:dyn} which captures and illustrates the main novelties
of fractional types and their runtime representations as first-class
gc processes. This definition allows allocation and de-allocation of
ancillae in patterns beyond the scope model but, like existing
stack-based solutions, still requires a runtime check to verify the
safety of de-allocation.
%% \ednote{JC: I don't understand the previous sentence. What does
%% ``this presentation'' refer to?  The paper as a whole, or the
%% details of the fractional types? There is also a tension between
%% ``partially solves'' and ``more general and more expressive'' which
%% might very well be true, but makes for a very vague sentence.}
To make the arguments regarding memory
management explicit, Sec.~\ref{sec:space} presents an abstract machine
with a heap component that grows and shrinks as ancillae are allocated
and de-allocated via garbage collection. In Sec.~\ref{sec:dep} we
lift ``plain'' programs manipulating ancillae to the type system in
order to exploit the power of dependent types to guarantee, for the
first time, the condition necessary for safe de-allocation of
ancillae. Technically this solution requires the introduction of
pointed types on top of which are defined both a monad and comonad of
singleton types. The tracking of ancillae value is done within the
type system via the singleton types. A number of examples exploiting
the singleton-based type system are presented to demonstrate the full
expressiveness of reasoning about fractional types. The type safety proofs demonstrate
that plain (but guaranteed safe)
programs can be extracted and executed without errors or runtime
checks in the conventional type system. The last section concludes
with a discussion of related work and a summary of our results.

@@

\begin{center}
\input{revrev.tikz}
\end{center}

@@

\PPFTname{}

\PPFTtrace{}

\PPFThof{}
